---
title: "Untitled"
output: word_document
date: "2026-03-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Models

Starting with Anders' FSA model, we built out 2 versions each of the MA(1) and AR(1) models.

MA(1) models included

(1) Modeling the deviations $eps \sim N(0, SdR)$ ("basicfsa_ma.R")

(2) Modeling the deviations as $eps \sim N(0, 1)$ and scaling by SdR ("basicfsa_ma_z.R")

AR(1) models included

(1) Modeling the deviations as $\sim N(0, 1)$ and manually updating the values in a loop ("basicfsa_manual_ar.R")

(2) Implementing the sparse matrix version ("basicfsa_sparse_ar.R")

For each model, we generated 10 random datasets by corrupting the observations in Anders' initial data with additional lognormal observation error. We then did 2 types of MCMC sampling for each dataset:

(1) Passing RTMB models to `tmbstan`

(2) Fitting Models in Stan directly, skipping RTMB

For each model, we summarized

-   the maximum correlation across monitored parameters (`max_cor`)

-   the mean correlation across monitored parameters (`mean_cor`)

-   the time to run 4 MCMC chains with 3000 iterations (`sampling_time`, seconds)

-   the number of parameters with an Rhat statistic \> 1.15 (`pars_hi_rhat`, lower values better)

-   the maximum Rhat statistic (`max_r_hat`, lower values better)

-   the mean effective parameters sampled (`mean_n_eff`, higher values better)

## Results: MA models

Results from the MA models show that sampling with `tmbstan` is 10x faster than using Stan directly, all parameters converge quickly, and more effective samples are taken. Comparing the regular implementation to Z-scored deviations, we find that approximately half of the Z-scored models fail to pass convergence tests; as speed and efficiency are similar, there seems to be limited advantages of Z-scoring when combined with `tmbstan`.

## Results: MA(1) models with standard implementation

```{r echo = FALSE}
d <- readRDS("ma_results_basic.rds")
names(d) <- c("i", "Max(cor)", "Mean(cor)", "Time", "N_hi_Rhat", "Max(Rhat)","N_eff", "Type")
knitr::kable(d, digits = 2)

```

## Results: MA(1) models with Z-score implementation
```{r echo = FALSE}
d <- readRDS("ma_results_z.rds")
names(d) <- c("i", "Max(cor)", "Mean(cor)", "Time", "N_hi_Rhat", "Max(Rhat)","N_eff", "Type")
knitr::kable(d, digits = 2)

```

## Results: AR(1) models with manual implementation

```{r echo = FALSE}
d <- readRDS("ar_results_manual.rds")
names(d) <- c("i", "Max(cor)", "Mean(cor)", "Time", "N_hi_Rhat", "Max(Rhat)","N_eff", "Type")
knitr::kable(d, digits = 2)
```

## Results: AR(1) models with sparse implementation

```{r echo = FALSE}
d <- readRDS("ar_results_sparse.rds")
names(d) <- c("i", "Max(cor)", "Mean(cor)", "Time", "N_hi_Rhat", "Max(Rhat)","N_eff")
d$Type <- "tmbstan"
knitr::kable(d, digits = 2)
```
