---
title: ''
author: ''
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(knitr)
```

## Abstract

---

## Introduction

---

## Methods

### Model structure

We used an age-structured model based on the `simplefsa` fisheries dataset [more detail from Anders]. The model tracks log-numbers-at-age ($\log N_{a,y}$) over age classes $a = 1, \ldots, A$ and years $y = 1, \ldots, T$. Fishing mortality is modeled as a separable age-year product $F_{a,y} = \exp(\log F_a + \log F_y)$. The population is propagated forward by

$$\log N_{a,y} = \log N_{a-1,y-1} - F_{a-1,y-1} - M_{a-1,y-1}$$

where $M_{a,y}$ is natural mortality supplied as data. Observations consist of commercial catch-at-age (fleet 1) and survey indices (fleet 2, ages 1–5), both modeled with lognormal observation error.

**Recruitment dynamics.** Two formulations were implemented:

#### MA(1) recruitment 

Annual recruitment deviations $\varepsilon_y$ follow a first-order moving-average process,

$$\log N_{1,y} = \mu_R + \varepsilon_y + \theta \varepsilon_{y-1}, \quad \varepsilon_y \sim N(0, \sigma_R)$$

where $\theta = 2\,\text{logit}^{-1}(\tilde\theta_R) - 1$.

#### AR(1) recruitment

Recruitment follows a first order autoregressive process on log-recruitment,

$$\log N_{1,y} = \phi \log N_{1,y-1} + \mu_R(1 - \phi) + \sigma_R z_y, \quad z_y \sim N(0,1)$$
where $\phi = 2\,\text{logit}^{-1}(\tilde\phi_R) - 1$.

**Parameterisation.** Two implementations were compared for each recruitment model:

For the MA(1) model we compared 
- *Standard* (MA only): random effect $\varepsilon \sim N(0, \sigma_R)$ (centred)
- *Non-centred* (both): random effect $z \sim N(0,1)$, scaled by $\sigma_R$ inside the likelihood

For the AR(1) model we compared 
- *Sparse*: uses RTMB/TMB's `dautoreg()` function
- *Non-centred*: random effect $z \sim N(0,1)$, scaled by $\sigma_R$ inside the likelihood

#### Priors 

All standard deviations ($\sigma_\text{catch}$, $\sigma_\text{survey}$, $\sigma_R$) were given $N(0,1)$ priors on the log scale. All other parameters were given flat priors.

### MCMC sampling

For each model, 100 replicate datasets were generated by perturbing observations from the base dataset. Each dataset was fit using two MCMC approaches:

1. tmbstan: the fitted RTMB model object was passed directly to Stan's MCMC sampler

2. Stan: the model was coded directly in Stan and sampled with MCMC

Both approaches used 4 chains, 4000 iterations (2000 warmup), and `adapt_delta = 0.99`. Chains were run sequentially due to RTMB namespace constraints in worker processes.

### Summary statistics

For each model $\times$ implementation combination we summarised across the 100 replicates: mean sampling time (seconds), mean and maximum $\hat{R}$, mean effective sample size (ESS), and the number of parameters with $\hat{R} > 1.15$. Results are presented as mean (SD) across replicates.

---

## Results

```{r load-data}
## Load result files and standardize column names
load_results <- function(path, model_label) {
  d <- readRDS(path)
  
  if ("sampling" %in% names(d)) {
    names(d)[names(d) == "sampling"] <- "Type"
  }
  col_map <- c(
    max_cor       = "max_cor",
    mean_cor      = "mean_cor",
    sampling_time = "sampling_time",
    pars_hi_rhat  = "pars_hi_rhat",
    max_r_hat     = "max_r_hat",
    mean_n_eff    = "mean_n_eff",
    time_s        = "sampling_time",
    max_rhat      = "max_r_hat",
    mean_neff     = "mean_n_eff",
    min_neff      = "min_n_eff"
  )
  for (old in names(col_map)) {
    if (old %in% names(d)) names(d)[names(d) == old] <- col_map[old]
  }
  d$model <- model_label
  d
}

ma_centered    <- load_results("ma_centered_results.rds",    "MA(1) centred")
ma_noncentered <- load_results("ma_noncentered_results.rds", "MA(1) non-centred")
ar1            <- load_results("ar1_results.rds",            "AR(1)")

all_results <- bind_rows(ma_centered, ma_noncentered, ar1)
all_results$Type  <- factor(all_results$method,  levels = c("tmbstan", "stan"))
all_results$model <- factor(all_results$model,
                            levels = c("MA(1) centred", "MA(1) non-centred", "AR(1)"))
```

```{r summary-fn}
## Summarise key metrics across replicates: mean (SD)
fmt_mean_sd <- function(x) {
  x <- x[is.finite(x)]
  sprintf("%.2f (%.2f)", mean(x), sd(x))
}

make_summary_table <- function(df) {
  metrics <- c("sampling_time", "max_r_hat", "mean_n_eff", "pars_hi_rhat")
  labels  <- c("Sampling time (s)", "Max Rhat", "Mean ESS",
               "N pars Rhat > 1.15")

  types <- unique(as.character(df$method))
  out   <- data.frame(Metric = labels)

  for (tp in types) {
    sub <- df[df$method == tp, ]
    col <- sapply(metrics, function(m) {
      if (m %in% names(sub)) fmt_mean_sd(sub[[m]]) else "—"
    })
    out[[tp]] <- col
  }
  out
}
```

### MA(1) models

```{r table-ma-basic, results='asis'}
tab <- make_summary_table(ma_centered)
kable(tab, caption = "Table 1. MA(1) centred implementation. Mean (SD) across 100 replicates.",
      align = c("l", "r", "r"))
```

```{r table-ma-z, results='asis'}
tab <- make_summary_table(ma_noncentered)
kable(tab, caption = "Table 2. MA(1) non-centred implementation. Mean (SD) across 100 replicates.",
      align = c("l", "r", "r"))
```

### AR(1) models

Results for the AR(1) model are presented in Table 3.

```{r table-ar-manual, results='asis'}
tab <- make_summary_table(ar1)
kable(tab, caption = "Table 3. AR(1) implementation. Mean (SD) across 100 replicates.",
      align = c("l", "r", "r"))
```

### Results

#### Sampling time

- tmbstan and Stan have nearly identical times for the MA(1) centered model, but tmbstan is faster for the MA(1) non-centered model (~83s vs ~118s) and the AR(1) model (~391s vs ~482s).

- The AR(1) model is ~4–5× slower than either MA(1) variant for both samplers

#### R-hat

- For MA(1) centred, both samplers converge similarly, with no strong advantage for either approach.

For MA(1) non-centered, tmbstan converges considerably better than Stan, and Stan shows higher run-to-run variability. 

The AR(1) model shows poor convergence for both samplers, suggesting we need to run the chains longer

#### ESS

- As expected, Stan consistently achieves higher mean ESS across all three models

#### Centered v non-centered?

- For tmbstan, the two parameterisations perform nearly identically in all metrics

- For Stan, the non-centred version is ~30% slower and produces substantially worse convergence
