---
title: 'Working Title - Sparsity in random effects under different parameterizations in fishery stock assessment models'
author: 'Andrea M Havron$^{a,1,*}$, Anders Nielsen$^{b,2}$, Nathan Vaughan$^{c,3}$, Eric J. Ward$^{d,4}$'
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  bookdown::word_document2:
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(knitr)
```

\
$^{a}$NOAA Fisheries, Office of Science and Technology, Silver Spring, MD, USA  
$^{b}$Danish Technical University, Aqua, Kongens Lyngby, Denmark  
$^{c}$Vaughan Analytics, in support of NOAA Fisheries Southeast Fisheries Science Center, Miami, FL, USA  
$^{d}$NOAA Fisheries, Northwest Fisheries Science Center, Seattle, WA, USA  
\
\
$^{*}$**Corresponding author at:**  
NOAA Fisheries, Office of Science and Technology, 1315 East-West Highway, 12th Floor, Silver Spring, MD, 20190, USA  
Email address: andrea.havron@noaa.gov  
\
\
ORCID ID:  
Andrea Havron: 0000-0002-4080-448X  
Anders Nielsen: 0000-0001-9683-9262  
Nathan Vaughan: 0009-0000-3054-6950  
Eric Ward: 0000-0002-4359-0296  

\newpage

# Abstract{-}

  **[250 word limit]:**   
  Many stock assessment models in the U.S. have been developed in ADMB using a hierarchical approach for time-varying processes, where mean parameter values $\bar{\theta}$ and fixed effect annual deviation values $dev_{y}$ are estimated and combined to produce annual parameter estimates $\hat{\theta}_{y} = \bar{\theta} * dev_{y}$. This structure was intuitive and enabled parameter means to be estimated in earlier phases than deviations, which improved convergence. Following statistical best practices, several next generation stock assessment models have incorporated the use of random effects through TMB to model process variability in time-varying parameters. This approach does not require parameter phasing but does require the calculation of up to third-order partial derivatives and Hessian inversion at each optimization step to integrate out random effect uncertainty using the Laplace approximation. While simple to interpret and implement, the deviations approach results in a dense covariance matrix of correlations 
  between the mean $\bar{\theta}$ and each annual deviation $dev_{y}$ estimate. Due to this difference, the deviations approach in a random effects model has a much larger computational penalty than in historic fixed effects modeling platforms.  We test an alternate configuration of directly estimating annual recruitment values as random effects in AR1 and MA1 processes. While the former structure has a sparse covariance matrix and reduces model estimation time, the latter produces the opposite result. Optimization time and parameter estimates from stock assessment models under these equivalent sparse versus dense parameterizations are compared to confirm equivalency and performance improvement of the new structure. These results inform development in next generation stock assessment models.

# Keywords{-} 
You are required to provide 1 to 7 keywords for indexing purposes. Keywords should be written in English. Please try to avoid keywords consisting of multiple words (using "and" or "of").

We recommend that you only use abbreviations in keywords if they are firmly established in the field.

# Highlights{-}
You are required to provide article highlights at submission.

Highlights are a short collection of bullet points that should capture the novel results of your research as well as any new methods used during your study. Highlights will help increase the discoverability of your article via search engines. Some guidelines:

- Submit highlights as a separate editable file in the online submission system with the word "highlights" included in the file name.

- Highlights should consist of 3 to 5 bullet points, each a maximum of 85 characters, including spaces.

\newpage
# Notes for Manuscript Development
Key points:

-   Lots of variable implementations of random effects for recruitment - deviations vs recruitment, often chosen for convenience

-   Unclear whether ML results will translate to Bayesian estimation

    -   lots of history in comparing centered vs non-centered REs

    -   best practices in creating sparsity pretty well understood in TMB, less clear what this means for MCMC

Gap / objective - tools such as tmbstan now allow TMB/RTMB models to be
passed directly to Stan

-   meaningful comparisons require verifying that the RTMB and Stan implementations encode the same posterior

-   we systematically compare sampling efficiency, convergence, and effective sample size across five combinations of MA(1) and AR(1) recruitment models, using both RTMB-tmbstan and Stan
    1. Is treating recruitment deviations or recruitment more efficient?
    2. What is the impact of RTMB's dautoreg function?

-   other?

* need to decide on jargon: innovations vs deviations? process v. log-recruitment? etc

# Introduction
Drawing from the Bayesian literature, it is clear that model efficiency is not merely a matter of software implementation but is fundamentally tied to the geometric properties of the posterior distribution. In MCMC sampling, particularly within the Hamiltonian Monte Carlo (HMC) framework utilized by Stan, the choice between centered and non-centered parameterizations is often a response to the funnel geometry pathology (Betancourt & Girolami, 2015). This pathology occurs when a sampler struggles to explore the narrow regions of a distribution where the group-level variance is small -- a common issue in hierarchical models with sparse data (Monnahan et al., 2016). Specifically, a non-centered parameterization effectively stretches this funnel, allowing HMC to sample more efficiently by breaking the strong a priori dependence between hierarchical levels (Gorinova et al., 2019). However, if the data are highly informative, a centered parameterization may actually be more efficient, as the data can "pin down" the parameters and reduce the correlations that otherwise slow down the sampler (Papaspiliopoulos et al., 2007; Gorinova et al., 2019). 

Within the context of fisheries stock assessment models, there are many examples of hierarchical variables being estimated with centered or non-centered pararmeterizations; one of the clearest example of this is when variability is added to recruitment either by modeling the process itself, $recruitment ~ N(\mu_{r}, \sigma_{r})$ (centered parameterization), or by modeling the deviations around a mean, $deviations ~ N(0, \sigma_{r}), recruitment = \mu_{r}*\sigma_{r}$ (non-centered parameterizations). Historically, models reliant on ADMB estimate the deviations as fixed effects undered a constrained optimization where the variance, $\sigma_{r}$ is fixed. In several tools, such as Stock Synthesis, sum to zero constraints have been applied to ensure that Hessian matrices are positive definite. A fixed-effect vector of deviation parameters does not require second order partial derivatives to be estimated until model optimization is complete and variance is estimated.

Several next generation stock assessment models have changed dependency from ADMB to Template Model Builder (Kristensen et al., 2016) in order to incorporate the use of the Laplace approximation to integrate out hierarchical variables (i.e. random effects), allowing the estimation of the hierarchical variance parameter, e.g. $\sigma_{r}$ (SAM, WHAM, FIMS). The Laplace approximation requires up to third-order derivates for its evaluations, thus necessitating a Jacobian calculation and Hessian inversion at each optimization step.  

To quantify the magnitude of this penalty we investigate the performance and optimization time of a simple random effects model under equivalent sparse vs dense parameterizations. For this example we utilize two simple proxy models; an auto regressive AR1 function and a simple stock assessment model based on babySAM. These proxies should reflect the relative performance gains expected in the real-world correlated time-varying processes modeled in stock assessments such as recruitment. Each of these time-series models can be parameterized as a dense matrix, representing the traditional parameter deviations as a random-effects approach, or as a sparse matrix, as would be achieved by directly estimating annual parameter values as random effects.

These two approaches are anticipated to produce mathematically equivalent results while differing in their computational overhead. Confirming this equivalency and quantifying the runtime improvement will provide support for the change in approach relative to previous assessment models.

Betancourt, M., & Girolami, M. (2015). https://arxiv.org/abs/1312.0906 

Gorinova 2019 https://doi.org/10.48550/arxiv.1906.03028

Papaspiliopoulos 2007 https://projecteuclid.org/journals/statistical-science/volume-22/issue-1/A-General-Framework-for-the-Parametrization-of-Hierarchical-Models/10.1214/088342307000000014.full

Kristensen et al. (2016) https:://doi.org/10.18637/jss.v070.i05

EW: Could add ecological refs

# Methods

## Simulation models

We used an age-structured model based on the `simplefsa` fisheries
dataset [more detail from Anders]. The model tracks log-numbers-at-age
($\log N_{a,y}$) over age classes $a = 1, \ldots, A$ and years
$y = 1, \ldots, T$. Fishing mortality is modeled as a separable age-year
product $F_{a,y} = \exp(\log F_a + \log F_y)$. The population is
propagated forward by

$$\log N_{a,y} = \log N_{a-1,y-1} - F_{a-1,y-1} - M_{a-1,y-1}$$

where $M_{a,y}$ is natural mortality supplied as data. Observations
consist of commercial catch-at-age (fleet 1) and survey indices (fleet
2, ages 1–5), both modeled with lognormal observation error.

We implemented two formulations of recruitment dynamics, not in the same
model. First, we modeled annual recruitment deviations $\varepsilon_y$
as a first-order moving-average process (MA(1)),

$$\log N_{1,y} = \mu_R + \varepsilon_y + \theta \varepsilon_{y-1}, \quad \varepsilon_y \sim N(0, \sigma_R)$$

where $\theta = 2\,\text{logit}^{-1}(\tilde\theta_R) - 1$.

Our second formulation simulated annual recruitment as a first order
autoregressive AR(1) process,

$$\log N_{1,y} = \phi \log N_{1,y-1} + \mu_R(1 - \phi) + \sigma_R z_y, \quad z_y \sim N(0,1)$$
where the AR parameter $\phi = 2\,\text{logit}^{-1}(\tilde\phi_R) - 1$.

## Estimation

We compared five model implementations across the two recruitment
dynamics formulations (MA(1) and AR(1)), varying in the choice of random
effect and the resulting Hessian sparsity structure (Table \@ref(tab:table1)). For the
MA(1) models, we compared treating the innovations $\varepsilon_y$ or
recruitment $\log N_{1,y}$ directly as the random effect. When
innovations $\varepsilon_y$ are a random effect, $\log N_{1,y}$ is
computed deterministically MA(1) process, yielding a sparse banded
Hessian because each $\varepsilon_y$ connects only to its immediate
neighbors. In our second MA(1) estimation model with $\log N_{1,y}$ as
random effects, the $\varepsilon_y$ become derived quantities, with a
dense Hessian because each $\log N_{1,y}$ propagates through all future
time steps. [cite Kristensen 2016? Rue and Held?, other?]

For the AR(1) models, we implemented three estimation model variants.
RTMB includes a `dautoreg` function, which uses the banded precision
matrix of the AR(1) process for sparse Hessian evaluation. Our first
estimation approach treated the $\log N_{1,y}$ as random effects, and
utilized the `dautoreg` approach. To isolate the effect of this
function, our second approach involved an identical model, but we
replaced `dautoreg` with a loop, resulting in a dense Hessian from the
same statistical model. Parallel to our MA(1) comparison, we were also
interested in effects of treating the deviations as random effects; our
third estimation model treated the AR(1) deviations
$\varepsilon_y \sim N(0,\, \sigma_R\sqrt{1-\phi^2})$ as random effects,
with $\log N_{1,y}$ as derived quantities, also yielding a sparse
Hessian. Collectively these implementations allow us to disentangle the
effects of random effect choice and Hessian sparsity on sampling
efficiency.

We performed estimation using maximum likelihood and Bayesian approaches
(MCMC). Models were coded in RTMB (cite), and fitted model objects were
then passed directly to Stan via the `tmbstan` package (cite Monahan et
al.). For comparison purposes, we then re-coded estimation models
directly in Stan, and re-estimated parameters. Both MCMC approaches used
4 chains, with 4000 iterations each (retaining the last 2000). We
specified the target acceptance probability of the NUTS sampler to be
0.99 (`adapt_delta`). For Bayesian estimation (and RTMB equivalence),
all standard deviation parameters were given $N(0,1)$ priors on the log
scale, and all other parameters were given Stan's default flat priors.

```{r table1, echo=FALSE}
model_table <- data.frame(
  Model = c(
    "MA",
    "MA",
    "AR",
    "AR",
    "AR"
  ),
  `Random effect` = c(
    "$\\varepsilon$",
    "$\\log N_{1,y}$",
    "$\\log N_{1,y}$ (dautoreg)",
    "$\\log N_{1,y}$ (loop)",
    "$\\varepsilon$"
  ),
  Hessian = c("Sparse", "Dense", "Sparse", "Dense", "Sparse"),
  check.names = FALSE
)

knitr::kable(model_table,
  caption = "Summary of model implementations.",
  align   = c("l", "l", "c"),
  escape  = FALSE
)
```

## Summary statistics

For each combination of estimation model and implementation (RTMB, tmbstan, Stan) we summarized across the 100 replicates: mean sampling time
(seconds), mean and maximum $\hat{R}$, mean effective sample size (ESS),
and the number of parameters with $\hat{R} > 1.15$. Results are
presented as mean (SD) across replicates.

# Results

For the AR(1) formulation, we found the strongest support for implementing random effects using the process (log-recruitment) as random effects, instead of the innovations (deviations). Across the 100 simulated datasets, the maximum likelihood optimization in RTMB was faster on average (Table \@ref(tab:table-ar-nodautoreg) vs Table \@ref(table-ar-deviations)). This difference becomes even stronger with our Bayesian comparisons. While we found similar effective sample size (ESS) and run time between `tmbstan` and `rstan`, parameterizing random effects in log-recruitment was more efficient: this approach resulted in 105.1 and 78.6 samples / second for `tmbstan` and `rstan` respectively, versus 6.4 and 17.5 samples / second for each when parameterizing random effects in innovations (Table \@ref(tab:table-ar-nodautoreg) vs Table \@ref(table-ar-deviations)). We found little difference in the AR(1) formulation on maximum likelihood estimation with and without RTMB's `dautoreg` function, (Table \@ref(tab:table-ar-dautoreg) vs Table \@ref(tab:table-ar-nodautoreg)).

For the MA(1) estimation model, we found that implementing random effects in the deviations resulted in faster maximum likelihood optimization (0.09 sec vs 0.15 sec, \@ref(tab:table-ma-good) \@ref(tab:table-ma-bad) ). Stan and tmbstan implementations had similar average speeds across the two implementations with slight efficiency advantages in `tmbstan` -- with random effects in deviations, `tmbstan` and `rstan` sampled 32.7 and 23.9 samples / sec, respectively, while with random effects in the process (log-recruitment), these approaches samples 30.57 and 25.2 samples / sec (0.09 sec vs 0.15 sec, \@ref(tab:table-ma-good) \@ref(tab:table-ma-bad) ).


```{r load-data}
## Load result files and standardize column names
load_results <- function(path, model_label) {
  d <- readRDS(path)

  if ("sampling" %in% names(d)) {
    names(d)[names(d) == "sampling"] <- "Type"
  }
  col_map <- c(
    sampling_time = "sampling_time",
    pars_hi_rhat  = "pars_hi_rhat",
    max_r_hat     = "max_r_hat",
    mean_n_eff    = "mean_n_eff",
    time_s        = "sampling_time",
    max_rhat      = "max_r_hat",
    mean_neff     = "mean_n_eff",
    min_neff      = "min_n_eff"
  )
  for (old in names(col_map)) {
    if (old %in% names(d)) names(d)[names(d) == old] <- col_map[old]
  }
  d$model <- model_label
  d
}

ma_good <- load_results(path = "../bayes/ma_good_results.rds", "MA(1) good")
ma_bad <- load_results(path = "../bayes/ma_bad_results.rds", "MA(1) bad")
# ar1 <- load_results("ar1_results.rds", "AR(1)")

ma_results <- bind_rows(ma_good, ma_bad)
ma_results$method[which(ma_results$method == "tmb")] <- "RTMB"
ma_results$Type <- factor(ma_results$method, levels = c("stan", "tmbstan", "RTMB"))

ar_good <- load_results(path = "../bayes/ar_good_results.rds", "AR(1) good")
ar_good_eps <- load_results(path = "../bayes/ar_good_eps_results.rds", "AR(1) eps")
ar_good_nodautoreg <- load_results(path = "../bayes/ar_good_nodautoreg_results.rds", "AR(1) no dautoreg")
```

```{r summary-fn}
## Summarise key metrics across replicates: mean (SD)
fmt_mean_sd <- function(x) {
  x <- x[is.finite(x)]
  sprintf("%.2f (%.2f)", mean(x), sd(x))
}

make_summary_table <- function(df) {
  metrics <- c(
    "sampling_time",
    # "max_r_hat",
    "mean_n_eff",
    "pars_hi_rhat"
  )
  labels <- c(
    "Sampling time (s)",
    # "Max Rhat",
    "Mean ESS",
    "N pars Rhat > 1.15"
  )

  types <- unique(as.character(df$method))
  out <- data.frame(Metric = labels)

  for (tp in types) {
    sub <- df[df$method == tp, ]
    col <- sapply(metrics, function(m) {
      if (m %in% names(sub)) fmt_mean_sd(sub[[m]]) else "—"
    })
    out[[tp]] <- col
  }

  out <- rbind(
    out,
    data.frame(
      "Metric" = "All Rhat < 1.15",
      tmb = NA,
      tmbstan = length(which(df$method == "tmbstan" & df$max_r_hat < 1.15)),
      stan = length(which(df$method == "stan" & df$max_r_hat < 1.15))
    )
  )

  out
}
```

```{r direct-comparison, echo=FALSE}
ma_good_rtmb <- dplyr::filter(ma_good, method == "tmb")
ma_bad_rtmb <- dplyr::filter(ma_bad, method == "tmb")
ma_good_stan <- dplyr::filter(ma_good, method == "stan")
ma_bad_stan <- dplyr::filter(ma_bad, method == "stan")
ma_good_tmbstan <- dplyr::filter(ma_good, method == "tmbstan")
ma_bad_tmbstan <- dplyr::filter(ma_bad, method == "tmbstan")
```


```{r table-ar-nodautoreg, results='asis'}
tab <- make_summary_table(ar_good_nodautoreg)
kable(tab,
  caption = "AR(1) implementation without dautoreg, and log-recruitment as random effects. Mean (SD) across 100 replicates.",
  align = c("l", "r", "r", "r")
)
```

```{r table-ar-deviations, results='asis'}
tab <- make_summary_table(ar_good_eps)
kable(tab,
  caption = "AR(1) implementation without dautoreg, and deviations as random effects. Mean (SD) across 100 replicates.",
  align = c("l", "r", "r", "r")
)
```

```{r table-ar-dautoreg, results='asis'}
tab <- make_summary_table(ar_good)
kable(tab,
  caption = "AR(1) implementation with dautoreg, and log-recruitment as random effects. Mean (SD) across 100 replicates. Because Stan doesn't have a dautoreg function, we compare against the Stan code without dautoreg.",
  align = c("l", "r", "r", "r")
)
```

```{r table-ma-good, results='asis'}
tab <- make_summary_table(ma_good)
kable(tab,
  caption = "MA(1) implementation, with deviations as random effects. Mean (SD) across 100 replicates.",
  align = c("l", "r", "r", "r")
)
```

```{r table-ma-bad, results='asis'}
tab <- make_summary_table(ma_bad)
kable(tab,
  caption = "MA(1) implementation, with log-recruitment as random effects. Mean (SD) across 100 replicates.",
  align = c("l", "r", "r", "r")
)
```

# Discussion

- For MA(1) models, we recommend modelling deviations as random effects (good parameterisation). The RTMB implementation was much faster, and treating recruitment as random effects led to really poor estimation for some Bayesian models (Stan)

- For AR(1) models, we recommend using recruitment as random effects. This was much faster than the deviations parameterization

- RTMB's 'dautoreg' function seemed to have little to no impact on MCMC sampling speed or efficiency
